---
title: "Homework #1"
author: "Suruchi Ahuja, Abbas Rizvi, Hoi Lam Tai, Jingchen Zhang"
date: "February 15, 2016"
output: pdf_document
---
# Problem 1
Dr. Gaile provided  `R` code to explore 'potentially interesting genes'. Important portions of Dr. Gaile's code will be pasted in `R` chunks throughout this submission. 

The dataset `gse19439.RData` from Lecture 04 (2016) was loaded.
```{r, message=FALSE}
load("/Users/aarizvi/Dropbox/Group2/HW1/abbas_hw1/gse19439.RData")
gse19439
```

`gse19439` is an `ExpressionSet` object. An `ExpressionSet` contains gene expression data from a microarray experiment. The `ExpressionSet` also contains 'meta-data', which can be accessed using functions such as `phenoData`, which describes the samples in the experiment, or `featureData`, which contains probe IDs and other descriptive information. For more information on `ExpressionSet` objects please see the Bioconductor vignette  [ExpressionSetIntroduction](http://bioconductor.org/packages/3.3/bioc/vignettes/Biobase/inst/doc/ExpressionSetIntroduction.pdf). 


To access the expression dataset that is contained within the `ExpressionSet`, the function `exprs()` can be utilized. We assigned `X` as our expression data. 
```{r}
X <- exprs(gse19439)
```

A Kruskal Wallis scan was conducted in order to test whether the samples originated from the same distribution. 

```{r, eval=FALSE}
myKrusk <- function(i){
        cat(i,"...",fill=F)
        kruskal.test(x=X[i,], g=FTB)$p.value
}
```

The p-values that were calculated from the Kruskal-Wallis test were assigned as the numeric vector `myPvals`. The order of `myPvals` was subsequently rearranged in ascending order and the row position index was subsetted into `best4`. 

```{r}
load("/Users/aarizvi/Dropbox/Group2/HW1/abbas_hw1/myPvals.RData")
GroupLabels <- c("Group I","Group II","Group III","Group IV")
# pick the best 4 p-values and assign them to the students.
best4 <- order(myPvals)[1:4]
best4
```

Dr. Gaile assigned the order ranking (1 through 4) to STA 525 groups that were assigned the same value (e.g. `best4[2]` would be assigned to Group 2.)
Since we are group 2, our row assignment was determined to be `10685` (as shown above). We then subsetted the `gse19439` `ExpressionSet` to just contain information from our row using the variable `myrow`.

```{r}
myrow <- gse19439[10685,]
```

Our row corresponds to the probe ID ILMN_1703335.

```{r}
featureNames(myrow)
```
We accessed `featureData` of `myrow` to obtain additional information regarding the gene. The `featureData` revealed that the probe ID ILMN_1703335 corresponds to the 'LACTB' GeneSymbol. LACTB is an enzyme known as serine beta-lactamase-like protein. More information on LACTB is summarized in Table 1. 

```{r xtable,results="asis"}
gene.info <- pData(featureData(myrow))
gene.info <- t(gene.info)
library(xtable)
tab <- xtable(gene.info, caption="Group 2 Gene of Interest Summary")
print.xtable(tab,
             size = "footnotesize",
             include.rownames=TRUE,
             include.colnames=FALSE,
             comment=FALSE)
```

# Problem 2
The experimental conditions of the samples in `gse19439` belong to one of 3 different tuberculosis phenotype groups. The groups are: 1. control (CON), 2. latent TB (LTB), and 3. positive TB (PTB). The actual group that individual samples correspond to was stored in `phenoData` of `gse19439`. The groupings were accessed and trimmed into 3 letter abbreviations using `substring` for readability purposes. The groupings were assigned to the variable `FTB` which is of the class `factor`. This is very useful because it allows us to use `FTB` as an index when subsetting a `data.frame`.  

```{r}
# make a factor object for Control, latent TB and Positive TB
tmp <- as.character(pData(phenoData(gse19439))[,1])
J <- length(tmp) # J=number of samples
TBgroup <- rep("",J)
for(j in 1:J) TBgroup[j]=substring(tmp[j],1,3)
# make a factor for TBgroup
FTB <- factor(TBgroup,levels=c("CON","LTB","PTB"))
```

## 2.1 Evaluating LACTB Gene Expression Distribution per Phenotype Grouping
A boxplot, violin plot, and bean plot were produced in order to visualize the distribution of gene expression for LACTB. Each of the plots show the same general trend in regards to the groups. The mean of expression values for the LACTB gene is higher in PTB than LTB, and LTB than CON. The positive TB has the highest mean of expression values for LACTB, and the control group has the highest variances, which could have many different biological implications. In order to investigate these implications, a much more thorough study must be conducted with this gene and tuberculosis. An example biological implication could be that increased LACTB expression is a result from a positive TB infection and may potentially be a useful biomarker to differentiate from individuals whom possess the active disease from TB latent and TB negative individuals.

### 2.1.1 Boxplot
```{r}
boxplot(log2(X[10685,sampleNames(myrow)[FTB == 'CON']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'LTB']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'PTB']]),
        main = "Boxplot of Row Data as Function of TB Phenotype",
        names = c("CON", "LTB", "PTB"))
```

### 2.1.2 Violin Plot
```{r, message=FALSE}
require(vioplot)
vioplot(log2(X[10685,sampleNames(myrow)[FTB == 'CON']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'LTB']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'PTB']]),
        names = c("CON", "LTB", "PTB"))
title("Violin Plot of Row Data as Function of TB Phenotype")
```

### 2.1.3 Bean Plot
```{r, message=FALSE}
require(beanplot)
beanplot(log2(X[10685,sampleNames(myrow)[FTB == 'CON']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'LTB']]),
        log2(X[10685,sampleNames(myrow)[FTB == 'PTB']]),
        names = c("CON", "LTB", "PTB"))
title("Bean Plot of Row Data as Function of TB Phenotype")
```

# Problem 3
## 3.1 Heatmaps
A heatmap is a false color image with a dendrogram added to the left side and to the top. The best 20 student features were clustered and plotted in a heatmap using `heatmap.2`. The clustering method worked reasonably well such that members of the same group were closer in distance than members of other groups (see column labels of heatmap). 
```{r, message=FALSE}
require(gplots)
best20 <- order(myPvals)[1:20]
x.best <- X[best20,]
heatmap.2(X[best20,], main="20 student features", 
          scale = "row", trace="none", labCol = FTB)
```

As shown in the heatmap, the first 7 samples and the samples from 28 to 32 with label “PTB” are quite brighter than the others. This shows that the “PTB” group seems has a quite different distribution with the other two groups. The first 20 genes have higher expression frequency in “PTB” people than in other groups. However, the differences between group “CON” and “LTB” can not be that siginificant.

# Problem 4
The Affymetrix portion of the Platinum Spike dataset was considered. The `AffyBatch` object was stored as the variable `affydata`. The Affy Probe Spike Values were loaded into the data frame `spikeDF`. A comparative analysis of different ways for background correction, probe normalization, PM correction, summarization methods was conducted. We wanted to compare the effectiveness of eight different analysis routes. The final endpoint of our analysis was to generate ROC curves to be able to discriminate between the different analyses. The different routes are summarized in Table 2. The following subsections will describe our analysis workflow. 

## 4.1 Pre-processing
We first needed to pre-process the data, removing any NAs or 0s that are within the dataset.
```{r}
load("/Users/aarizvi/Dropbox/Group2/HW1/abbas_hw1/PSpikeAffyBatch.RData")
spikeDF <- read.table(file="/Users/aarizvi/Dropbox/Group2/HW1/abbas_hw1/AffyProbeSpikeValues.csv",sep="\t")
levels(spikeDF[,2])
summary(spikeDF[,2])

#grab Spike fold changes for all entries as numeric ...
#...so anything that was not a number is now an NA
SpikeFC <- as.numeric(levels(spikeDF[,2])[spikeDF[,2]])

#grab IDs
names(SpikeFC) <- spikeDF$V1

#remove NAs
nonZeroDX <- which(!is.na(SpikeFC) & (SpikeFC != 0)) 
spikeFC.clean <- SpikeFC[nonZeroDX]

#check how many genes are left in dataset
length(spikeFC.clean)
```

After the pre-processing, the dataset went from 18952 observations to 5370 observations. 

## 4.2 Normalization of 8 routes with expresso() and subset into ExpressionSet
The algorithms that were chosen were based off of their popularity and usage in the Halfon Spike Data. The algorithm of choice per method was added manually to a vector. A for loop was written to assign algorithms of choice in a vector and input them as arguments to the `expresso()` function from the Bioconductor `affy` package in an automated manner. 

```{r, results="asis"}
bgcorrect.mtd <- c("rma", "mas", "rma", "mas",
                   "rma", "none", "mas", "mas")
normalized.mtd <- c("constant", "quantiles", "quantiles", "loess",
                    "loess", "constant", "qspline", "qspline")
pmcorrect.mtd <- c("pmonly", "pmonly", "subtractmm", "mas",
                   "mas", "pmonly", "subtractmm", "mas")
summary.mtd <- c("mas", "mas", "avgdiff", "medianpolish", 
                 "medianpolish", "avgdiff", "mas", "mas")
route.df <- cbind(bgcorrect.mtd, normalized.mtd, pmcorrect.mtd, summary.mtd)
rownames(route.df) <- paste("Route", 1:8, sep = " ")
route.table <- xtable(route.df, caption = "Routes chosen for analysis")
print.xtable(route.table, size= "footnotesize", comment=FALSE)
```

```{r, eval=FALSE}
require(affy)
route.expsets <- list()
for (i in 1:length(bgcorrect.mtd)){
        routes <- expresso(affydata, 
                           bgcorrect.method = bgcorrect.mtd[i], 
                           normalize.method = normalized.mtd[i],
                           pmcorrect.method = pmcorrect.mtd[i], 
                           summary.method = summary.mtd[i])
        route.expsets[[i]] <- exprs(routes)[nonZeroDX,]
}
```


## 4.3 Multiple hypothesis testing
`mt.maxT` is a function used from the `multtest` library. This test is done to reduce the Type 1 and Type 2 errors. It computes permutation adjusted p-values for step-down maxP and minP multiple testing procedures.

```{r,eval=FALSE, message=FALSE}
#create labels for multitesting class labels -- 18 samples, 9 control, 9 experimental
labels <- factor(c(rep(0, 9), rep(1, 9))) #0 is control, 1 is experimental

require(multtest)
stats <- list()
for (i in 1:length(route.expsets)){
        #conduct multiple t test for 10000 permutations  
        testing.routes <- mt.maxT(route.expsets[[i]],
                                  classlabel = labels, B = 10000) 
        stats[[i]] <- testing.routes$adjp[order(testing.routes$index)]
}

nrow <- nrow(route.expsets[[1]])
myresponse <- rep(NA, nrow)
#if the spike value is 1 ... assign as 1s in matrix ... 
myresponse[which(spikeFC.clean == 1)] = 1 
#if spike value is not 1 ... assign as 0s in the matrix
myresponse[which(spikeFC.clean != 1)] = 0 
```

## 4.4 ROC Curves
Our ROC (receiver operating characteristic) curves are created by plotting the sensitivity against the specificity. So a plot with both higher sensitivity and higher specificity will be better. In our plot, the more a curve closes to the point in the upper left corner, the better the classifier will be. So among the 8 different methods we chose, the method with “mas/qspline/subtractmm/mas” (which AUC is 0.979) can be regarded as the best
one. 

```{r, eval=FALSE, message=FALSE}
#apply roc function on input 
roc.fnct <- function(x){roc(response = myresponse,
                            predictor = abs(x), plot=TRUE, print.auc=TRUE)}
#apply roc function on all the ordered test statistics in the list stats
roc <- lapply(stats, roc.fnct) 
```

```{r, message=FALSE}
load("/Users/aarizvi/Dropbox/Group2/HW1/abbas_hw1/question4.RData")
rainbow <-  palette(rainbow(length(route.expsets)))
plot(roc[[1]], main = "ROC curves for different normalization routes using expresso()", 
     col=rainbow[1])
legendText <- c()
for(i in 1:length(route.expsets)){
        plot(roc[[i]], add=TRUE, col=rainbow[i])
        legendText[i] <- paste(bgcorrect.mtd[i],
                               "/",normalized.mtd[i],"/", 
                               pmcorrect.mtd[i],
                               "/",summary.mtd[i],
                               "   AUC: ",
                               round(as.numeric(roc[[i]]$auc),3), sep="")
}
legend("bottomright",
       legendText,
       lty=c(rep(1,length(route.expsets))), 				
       lwd=c(rep(3,length(route.expsets))),
       col=rainbow,
       pt.cex=1,
       cex=0.75)
```


# Attribution of Work
Suruchi Ahuja - Involved with producing heatmaps and tables in R. Wrote section 4.3. 

Abbas Rizvi -  Wrote R scripts and sections 1-4 of report.

Hoi Lam Tai -  Involved with using `xtable` and learned from others R code  

Jingchen Zhang - Wrote R scripts and sections 1-4 report.
